Elaboration with First-Class Implicit Function Types

### 2020-06-02 paper.tex
revision

%% build: latexmk -pdf -pvc -xelatex paper.tex

@@

\newcommand{\yes}{\text{Yes}}
\newcommand{\yesst}{\text{Yes*}}
\newcommand{\no}{\text{No}}

@@

First, we present a variant of Norell's bidirectional elaborator
\cite[Chapter~3]{norell07thesis}. Compared to ibid.\ we make some extensions and
simplifications; what we end up with can be viewed as a toy version of the
actual Agda elaborator. Our main goal with the elaborator is to have a setup
which is as simple as possible, but sufficient to illustrate the enhancements in
Section \ref{sec:extending_elaboration}, and make it clear that this could be
used to conservatively extend existing dependently typed languages. Hence, we do
not have more fine-grained control over metavariable insertion, nor the advanced
unification algorithms used in Agda \cite{abel2011higher}; all of these are
orthogonal improvements to the presented system.

@@

Elaboration consists of two (partial) functions, checking and inferring, which
are defined by mutual induction on surface syntax. We also have implicit
argument insertion as a helper function, which is defined by recursion on core
types, and which is used as a post-processing step after inference. First, about
the used notations:
\begin{itemize}
  \item We use a Haskell-like monadic pseudocode notation, where the side effect is
    failure via $\fail$.
  \item We use pattern matching notation on core terms; e.g.\ we may match on
    whether a type is a function type. This assumes an evaluation/normalization
    procedure on core terms; but note that we already assume this in
    unification.
  \item We abbreviate $\theta_1 \circ \theta_2$ as $\theta_{12}$ and $\theta_1
    \circ \theta_2 \circ \theta_3$ as $\theta_{123}$ and analogously in other
    cases. We do this to reduce the visual noise caused by threading composed
    metasubstitutions everywhere in the elaboration algorithm.
\end{itemize}
We present the specifications and definitions below, then we describe them in order.
\begin{mathpar}
\inferrule*[lab=insert]{\Theta_0|\Gamma\vdash \\ (\Theta,\,\theta,\,t,\,A) \in
             \{(\Theta,\,\theta,\,t,\,A)\,|\,(\theta : \Theta \To \Theta_0)\,\land\,
               (\Theta|\Gamma[\theta]\vdash t : A)\}\cup\{\fail\}}
           {\einsert(\Theta,\,\theta,\,t,\,A) \in \{(\Theta,\,\theta,\,t,\,A)\,|\,
             (\theta : \Theta \To \Theta_0)\,\land\,(\Theta|\Gamma[\theta]\vdash t : A)\}\cup\{\fail\}}
\end{mathpar}
\begin{alignat*}{3}
  &\einsert\,(\Theta_1,\,\theta_1,\,t,\,\{x : A\}\to B) && :\equiv\hspace{0.5em}&&
       \slet\,(\Theta_2,\,\theta_2,\,u) = \freshMeta{\Theta_1}{\Gamma[\theta_1]}{A}\\
  & && &&\sin\,\einsert\,(\Theta_2,\,\theta_{12},\,(t[\theta_2])\,\{u\},\,B[\theta_2][x\mapsto u])\\
  &\einsert\,(\Theta_1,\,\theta_1,\,t,\,A) && :\equiv&& \ereturn\,(\Theta_1,\,\theta_1,\,t,\,A)\\
  &\einsert\,\fail && :\equiv && \fail
\end{alignat*}
\begingroup
\allowdisplaybreaks
\begin{mathpar}
\inferrule*[lab=check]
           {\text{$t$ is a surface expression} \\ \Theta|\Gamma \vdash A : \U}
           {\echeck{t}{\Theta}{\Gamma}{A} \in
             \{(\Theta',\,\theta,\,t')\,|\,(\theta : \Theta' \To \Theta)\,\land\,
               (\Theta'|\Gamma[\theta]\vdash t' : A[\theta])\}\cup\{\fail\}
           }

@@

  &\quad\eif\,(\Gamma = (\Gamma_0,\,x : A,\,\Gamma_1))\,\lor\,(\Gamma = (\Gamma_0,\,x : A = t,\,\Gamma_1)) \\

@@

  &\quad (\Theta_1,\,\theta_1,\,t',\,A) \leftarrow \einsert\,(\einfer{t}{\Theta_0}{\Gamma})\\

@@

  &\quad(\Theta_2,\,\theta_2,\,t',\,B)\leftarrow \eif\,t = \lambda\{y\}.\,u\,\ethen\,\einfer{t}{\Theta_1}{\Gamma[\theta_1],\,x:A}\\
  &\quad\quad\hspace{12em} \eelse\,\einsert(\einfer{t}{\Theta_1}{\Gamma[\theta_1],\,x:A}) \\

@@

  &\quad(\Theta_2,\,\theta_2,\,t',\,B)\leftarrow \eif\,t = \lambda\{y\}.\,u\,\ethen\,\einfer{t}{\Theta_1}{\Gamma[\theta_1],\,x:A}\\
  &\quad\quad\hspace{12em} \eelse\,\einsert(\einfer{t}{\Theta_1}{\Gamma[\theta_1],\,x:A}) \\

@@

The function $\einsert$ inserts implicit applications around a core term. For
example, if we have a defined name $id$ with type $\{A : \U\}\to A \to A$ in a
surface program, we usually want to expand $id$ to $id\,\{\alpha\}$, where
$\alpha$ is a fresh metavariable. We define $\einsert$ to take as input the
output of $\einferblank$, so that it is more convenient to use as an optional

@@

@@

The $\einfer{t\,u}{\Theta_0}{\Gamma}$ and $\einfer{t\,\{u\}}{\Theta_0}{\Gamma}$
cases are again interesting. Here, we first infer a type for $t$, then refine
the type to a function type, and lastly check the argument with the domain
type. Note the difference between the explicit and implicit case. In the former
case we use $\einsert\,(\einfer{t}{\Theta_0}{\Gamma})$, which inserts implicit
applications. In the latter case we do no insertion. This ensures that implicit
applications in the surface syntax behave similarly as in Agda. For example,
given $\mi{id} : \{A : \U\}\to A\to A$ in scope, we elaborate $\mi{id}\,\,\U$ as follows:
\begin{enumerate}
  \item The expression is an explicit application, so we infer $\mi{id}$ and insert implicit
        arguments, returning $\mi{id}\,\{\alpha\}$, where $\alpha$ is a fresh meta.
  \item We check that $\U$ has type $\alpha$. Here we immediately change direction,
        inferring $\U$ as type for $\U$ and unifying $\alpha$ with $\U$.
  \item Hence, the resulting output is $\mi{id}\,\{\U\}\,\U$.
\end{enumerate}
On the other hand, we elaborate $\mi{id}\,\{\U\}$ as follows:
\begin{enumerate}
  \item This is an implicit application, so we infer a type for $\mi{id}$ without inserting
    implicit arguments. This yields the inferred type $\{A : \U\}\to A\to A$, and
    we check that $\U$ has type $\U$.
  \item We change direction and infer $\U$ as type for $\U$, and successfully
    unify $\U$ with $\U$.
\end{enumerate}
It would be more efficient (and also allow more user-friendly error
messages) to not immediately refine the inferred type of $t$ to a function type,
but rather match on the inferred type, and only perform refining when the type
is meta-headed. We present the unoptimized version here for the sake of brevity.

For $\lambda$-s, we create a fresh meta for the domain type (since our surface
$\lambda$-s are not annotated), and infer types for the bodies. Note the extra
branching: we avoid insertion when the body is an implicit function, as in this
case this would yield a superfluous $\beta$-redex. We mainly do this in order to
match the insertion behavior of Agda, where the following definition works as
expected, without any over-eager insertion in the inner implicit $\lambda$.
\[
  \slet\,\mi{const} = \lambda\{A : \Set\}\{B : \Set\}(x : A)(y : B) \to x\,\sin\,...
\]
We do not have type-annotated $\lambda$-s in our system, but we can still
observe some examples which type-check with Agda-like insertion, and would fail
without the extra $\eif$ branch, like the following:
\begin{alignat*}{3}
  & \slet\,f : \_ = \lambda\,A\,\{B\}.\,A\,\sin\,\\
  & \slet\,g : \U \to \{B : \U\}\to U = f\,\sin\,...
\end{alignat*}
Note though that this is mostly a convenience feature in our prototype implementation,
which does support type-annotated $\lambda$-s.

@@

We remark that this notion of soundness is only a ``sanity'' or well-typing
statement for elaboration. In fact, we could define elaboration as a constantly
failing partial function, and it would also conform to the specification. The
right way to view this, is that $\echeckblank$ and $\einferblank$ together
with their specification constitute the semantics of surface syntax. We do not
give any other semantics to the surface syntax, nor does it support any other
operation.

@@

Why does this fail? The culprit is line $\scriptstyle{5}$, where we call
$\echeckt{\lambda\,x.\,x}{\alpha_0}$. At this point, the checking type is not an
implicit function type (it is a meta), so we do not insert an implicit
$\lambda$. At the heart of the issue is that elaboration makes insertion choices
based on core types.
\begin{enumerate}
\item $\echeck{t}{\Theta}{\Gamma}{A}$ can insert a $\lambda$ only if $A$ is an implicit function type.
\item $\einsert\,(\Theta,\,\theta,\,t,\,A)$ inserts an application only if $A$ is an implicit function type.
\end{enumerate}
In both of these cases, if $A$ is of the form $\alpha\,\overline{u}$
(i.e.\ meta-headed), then it is possible that $\alpha$ is later refined to an
implicit function, but at that point we have already missed our shot at implicit
insertion.

@@

\subsection{Unification For Strictly Curried Functions}
Although we omit most details of unification, we shall discuss it for curried
functions, as it is essential in the extended elaboration algorithm.  The most
interesting case is when we unify a curried function type with an implicit
function type. In this case, we learn that the domain of the curried function is
non-empty, so we refine the $A$ domain to an extended $(x_0 : A_0) \TCons A_1$
telescope. Since we invent a fresh $A_1$ domain for a curried function type, we
need to add a constancy constraint for it as well.
\begin{alignat*}{3}
& \unify_{\Theta_0|\Gamma}(\{x : \ol{A}\}\to B)\,(\{x_0 : A_0\}\to B') :\equiv \edo\\
& \quad \slet\,(\Theta_1,\,\theta_1,\,A_1) = \freshMeta{\Theta_0}{\Gamma,\,x_0 : A_0}{\Tel}\\
& \quad (\Theta_2,\,\theta_2) \leftarrow
  \unify_{\Theta_1|\Gamma[\theta_1]}\,(A[\theta_1])\,((x : A_0[\theta_1])\TCons A_1)\\
& \quad \slet\,(\Theta_3,\,\theta_3) =
    \newConstancy{\Theta_2}{\Gamma[\theta_{12}],\,x_0 : A_0[\theta_{12}]}{x_1}{(A_1[\theta_2])}{(B[\theta_{12}][x\mapsto (x_0 :: x_1)])}\\
& \quad\unify_{\Theta_3|\Gamma[\theta_{123}],\,x_0 : A_0[\theta_{123}]}\,
         (\{x_1 : A_1[\theta_{23}]\}\to B[\theta_{123}][x\mapsto (x_0 :: x_1)])\,B'
\end{alignat*}
We have the symmetric $\unify_{\Theta_0|\Gamma}(\{x_0 : A_0\}\to B')\,(\{x :
\ol{A}\}\to B)$ case the same way as above.

Now, let us assume that $B'$ is not an implicit function type, curried function
type or meta-headed. Then, we have the following case, where we solve a
telescope domain to be empty.
\begin{alignat*}{3}
& \unify_{\Theta_0|\Gamma}(\{x : \ol{A}\}\to B)\,B' :\equiv \edo\\
& \quad (\Theta_1,\,\theta_1) \leftarrow \unify_{\Theta_0|\Gamma}\,A\,\epsilon\\
& \quad \unify_{\Theta_1|\Gamma[\theta_1]}\,(B[\theta_1][x\mapsto[]])\,(B'[\theta_1])
\end{alignat*}
Again, we also have the symmetric case. For telescopes, $\Rec$, inhabitants of
$\Rec$, curried abstraction and application, unification is structural, and
other cases remain the same as in the basic elaborator of Section
\ref{sec:bidirectional_elaboration}.

@@

In the definition of checking, we insert a new clause after
$\echeck{t}{\Theta}{\Gamma}{(\{x : A\}\to B)}$:
\begin{alignat*}{3}
& \echeck{t}{\Theta_0}{\Gamma}{(\alpha\,\ol{u})} :\equiv \edo\\
& \quad \slet\,(\Theta_1,\,\theta_1,\,A) = \freshMeta{\Theta_0}{\Gamma}{\Tel}\\
&\quad(\Theta_2,\,\theta_2,\,t',\,B)\leftarrow \eif\,t = \lambda\{y\}.\,u\,\ethen\,\einfer{t}{\Theta_1}{\Gamma[\theta_1],\,x:\Rec\,A}\\
  &\quad\quad\hspace{12em} \eelse\,\einsert(\einfer{t}{\Theta_1}{\Gamma[\theta_1],\,x:\Rec\,A}) \\
& \quad \slet\,(\Theta_3,\,\theta_3) = \newConstancy{\Theta_2}{\Gamma[\theta_{12}]}{x}{(A[\theta_2])}{B}\\
& \quad (\Theta_4,\,\theta_4) \leftarrow
            \unify\, ((\alpha\,\ol{u})[\theta_{123}])\,(\{x : \ol{A[\theta_{23}]}\} \to B[\theta_3])\\
& \quad \ereturn\,(\Theta_4,\,\theta_{1234},\,(\lambda\,\{x : \ol{A[\theta_{234}]}\}.\,t'[\theta_{34}]))
\end{alignat*}
Hence, when checking a term with a meta-headed type, we create a fresh meta with
$\Tel$ type, infer a type for the term (avoiding superfluous insertion, as in
Section \ref{sec:elaboration} when inferring type for $\lambda$-s) and wrap the
result in a strictly curried $\lambda$. We again need to create new constancy
constraint. This way, if $B$ does not depend on $x : \Rec\,A$, the strictly
curried $\lambda$ in the output immediately computes away, since $A$ is solved
to $\epsilon$.

@@

On line $\scriptstyle{14}$ we have the case when a curried function type is
matched with a type which is not a curried function, implicit function or a
meta-headed type. We accordingly unify domain telescope $\alpha_3\,A$ with
$\epsilon$, which causes $\alpha_3$ to be solved as $\lambda\,A.\,\epsilon$
according to standard pattern unification. Now, $\alpha_2\,\{A\}\,\{z :
\ol{\alpha_3\,A}\}$ computes to $\alpha_2\,\{A\}\,\{z : \epsilon\}$, which
further computes to $\alpha_2\,\{A\}$ by \LabTirName{app-$\epsilon$}. From here,
unification finishes with success. On the last line, the expected output is
returned: since the $\alpha_1$ telescope meta is now solved, curried function
types and abstractions are computed away.
\end{example}

>>>>

%% build: latexmk -pdf -pvc paper.tex

@@

\newcommand{\yes}{\text{Yes}}
\newcommand{\yesst}{\text{Yes*}}
\newcommand{\no}{\text{No}}
\newcommand{\implArity}{\mathsf{implArity}}

@@

First, we present a variant of Norell's bidirectional elaborator
\cite[Chapter~3]{norell07thesis}. Compared to ibid.\ we make some extensions and
simplifications; what we end up with can be viewed as a toy version of the
actual Agda elaborator. Our main goal with the elaborator is to have a setup
which is as simple as possible, but sufficient to illustrate the enhancements in
Section \ref{sec:extending_elaboration}, and make it clear that this could be
used to conservatively extend elaboration in existing dependently typed
languages. Hence, we do not have more fine-grained control over metavariable
insertion, nor the advanced unification algorithms used in Agda
\cite{abel2011higher}; all of these are orthogonal improvements to the presented
system.

@@

Elaboration consists of two (partial) functions, checking and inferring, which
are defined by mutual induction on surface syntax. We also have implicit
argument insertion as a helper function, which is defined by recursion on core
types, and which is used as a post-processing step after inference. First, about
the used notations:
\begin{itemize}
  \item We use a Haskell-like monadic pseudocode notation, where the side effect is
    failure via $\fail$.
  \item We use pattern matching notation on core terms; e.g.\ we may match on
    whether a type is a function type. This assumes an evaluation/normalization
    procedure on core terms; but note that we already assume this in
    unification.
  \item We abbreviate $\theta_1 \circ \theta_2$ as $\theta_{12}$ and $\theta_1
    \circ \theta_2 \circ \theta_3$ as $\theta_{123}$ and analogously in other
    cases. We do this to reduce the visual noise caused by threading composed
    metasubstitutions everywhere in the elaboration algorithm.
\end{itemize}
We present the specifications and definitions below, then we describe them in order.
\begin{mathpar}
\inferrule*[lab=insertion,flushleft]{\Theta_0|\Gamma\vdash \\ (\Theta,\,\theta,\,t,\,A) \in
             \{(\Theta,\,\theta,\,t,\,A)\,|\,(\theta : \Theta \To \Theta_0)\,\land\,
               (\Theta|\Gamma[\theta]\vdash t : A)\}\cup\{\fail\}}
           {\einsert'(\Theta,\,\theta,\,t,\,A) \in \{(\Theta,\,\theta,\,t,\,A)\,|\,
             (\theta : \Theta \To \Theta_0)\,\land\,(\Theta|\Gamma[\theta]\vdash t : A)\}\cup\{\fail\}
           \\
            \einsert(\Theta,\,\theta,\,t,\,A)\hspace{0.3em}\in \{(\Theta,\,\theta,\,t,\,A)\,|\,
             (\theta : \Theta \To \Theta_0)\,\land\,(\Theta|\Gamma[\theta]\vdash t : A)\}\cup\{\fail\}
           }
\end{mathpar}
\begin{alignat*}{3}
  &\einsert'\,(\Theta_1,\,\theta_1,\,t,\,\{x : A\}\to B) && :\equiv\hspace{0.5em}&&
       \slet\,(\Theta_2,\,\theta_2,\,u) = \freshMeta{\Theta_1}{\Gamma[\theta_1]}{A}\\
  & && &&\sin\,\einsert'\,(\Theta_2,\,\theta_{12},\,(t[\theta_2])\,\{u\},\,B[\theta_2][x\mapsto u])\\
  &\einsert'\,(\Theta_1,\,\theta_1,\,t,\,A) && :\equiv&& \ereturn\,(\Theta_1,\,\theta_1,\,t,\,A)\\
  &\einsert'\,\fail && :\equiv && \fail\\
  &\einsert\,(\Theta,\,\theta,\,(\lambda\{x\}.t),\,A) && :\equiv&& \ereturn\,(\Theta,\,\theta,\,(\lambda\{x\}.t),\,A)\\
  &\einsert\,(\Theta,\,\theta,\,t,\,A) && :\equiv&& \einsert'\,(\Theta,\,\theta,\,t,\,A)\\
  &\einsert\,\fail && :\equiv && \fail\\
\end{alignat*}
\begingroup
\allowdisplaybreaks
\begin{mathpar}
\inferrule*[lab=check]
           {\text{$t$ is a surface expression} \\ \Theta|\Gamma \vdash A : \U}
           {\echeck{t}{\Theta}{\Gamma}{A} \in
             \{(\Theta',\,\theta,\,t')\,|\,(\theta : \Theta' \To \Theta)\,\land\,
               (\Theta'|\Gamma[\theta]\vdash t' : A[\theta])\}\cup\{\fail\}
           }

@@

  &\quad\eif\,(\Gamma \equiv (\Gamma_0,\,x : A,\,\Gamma_1))\,\lor\,(\Gamma \equiv (\Gamma_0,\,x : A = t,\,\Gamma_1)) \\

@@

  &\quad (\Theta_1,\,\theta_1,\,t',\,A) \leftarrow \einsert'\,(\einfer{t}{\Theta_0}{\Gamma})\\

@@

  &\quad(\Theta_2,\,\theta_2,\,t',\,B)\leftarrow \einsert\,(\einfer{t}{\Theta_1}{\Gamma[\theta_1],\,x:A})\\

@@

  &\quad(\Theta_2,\,\theta_2,\,t',\,B)\leftarrow \einsert\,(\einfer{t}{\Theta_1}{\Gamma[\theta_1],\,x:A}) \\

@@

We define two functions, $\einsert'$ and $\einsert$ which insert implicit
applications around a core term. For example, if we have a defined name $id$
with type $\{A : \U\}\to A \to A$ in a surface program, we usually want to
expand $id$ occurrences to $id\,\{\alpha\}$, where $\alpha$ is a fresh
metavariable. We define the insertion functions to take as input the output of
$\einferblank$, so that it is more convenient to use as an optional

@@

The difference between $\einsert$ and $\einsert'$ is that the latter avoids
insertion to implicit $\lambda$-terms. We usually want to avoid that, as it
creates a new implicit $\beta$-redex where the fresh metavariable is often
under-constrained. We only create such $\beta$-redex when the surface program
explicitly asks for it; we shall shortly see this in the discussion of the
$\einferblank$ case for explicit function application. With this, we largely
follow the insertion behavior of Agda.

@@

The $\einfer{t\,u}{\Theta_0}{\Gamma}$ and $\einfer{t\,\{u\}}{\Theta_0}{\Gamma}$
cases are again interesting. Here, we first infer a type for $t$, then refine
the type to a function type, and lastly check the argument with the domain
type. Note the difference between the explicit and implicit case. In the former
case we use $\einsert'\,(\einfer{t}{\Theta_0}{\Gamma})$, which inserts implicit
applications. In the latter case we do no insertion. This ensures that implicit
applications in the surface syntax behave similarly as in Agda. For example,
given $\mi{id} : \{A : \U\}\to A\to A$ in scope, we elaborate $\mi{id}\,\,\U$ as follows:
\begin{enumerate}
  \item The expression is an explicit application, so we infer $\mi{id}$ and insert implicit
        arguments, returning $\mi{id}\,\{\alpha\}$, where $\alpha$ is a fresh meta.
  \item We check that $\U$ has type $\alpha$. Here we immediately change direction,
        inferring $\U$ as type for $\U$ and unifying $\alpha$ with $\U$.
  \item Hence, the resulting output is $\mi{id}\,\{\U\}\,\U$.
\end{enumerate}
On the other hand, we elaborate $\mi{id}\,\{\U\}$ as follows:
\begin{enumerate}
  \item This is an implicit application, so we infer a type for $\mi{id}$ without inserting
    implicit arguments. This yields the inferred type $\{A : \U\}\to A\to A$, and
    we check that $\U$ has type $\U$.
  \item We change direction and infer $\U$ as type for $\U$, and successfully
    unify $\U$ with $\U$.
\end{enumerate}
Also note that we use $\einsert'$ here instead of $\einsert$. Thus, we get
expected (Agda-like) insertions with $\beta$-redexes as well. For example
$(\lambda\{A\}\,x.\, \mi{id}\,\{A\}\,x)\,\U$ is elaborated to
$(\lambda\{A\}\,x.\, \mi{id}\,\{A\}\,x)\,\{\U\}\,\U$.

It would be more efficient (and also allow more user-friendly error
messages) to not immediately refine the inferred type of $t$ to a function type,
but rather match on the inferred type, and only perform refining when the type
is meta-headed. We present the unoptimized version here for the sake of brevity.

For $\lambda$-s, we create a fresh meta for the domain type (since our surface
$\lambda$-s are not annotated), and infer types for the bodies.

@@

We remark that this notion of soundness is only a ``sanity'' or well-typing
statement for elaboration. In fact, we could define elaboration as a constantly
failing partial function, and it would also conform to the specification. One
way to view this is that $\echeckblank$ and $\einferblank$ together with their
specification constitute the semantics of surface syntax. We do not give any
other semantics to the surface syntax, nor does it support any other operation.

@@

Why does this fail? The culprit is line $\scriptstyle{5}$, where we call
$\echeckt{\lambda\,x.\,x}{\alpha_0}$. At this point, the checking type is not an
implicit function type (it is a meta), so we do not insert an implicit
$\lambda$. At the heart of the issue is that elaboration makes insertion choices
based on core types.
\begin{enumerate}
\item $\echeck{t}{\Theta}{\Gamma}{A}$ can insert a $\lambda$ only if $A$ is an implicit function type.
\item $\einsert\,(\Theta,\,\theta,\,t,\,A)$ can insert an application only if
  $A$ is an implicit function type.
\end{enumerate}
In both of these cases, if $A$ is of the form $\alpha\,\overline{u}$
(i.e.\ meta-headed), then it is possible that $\alpha$ is later refined to an
implicit function type, but at that point we have already missed our shot at
implicit insertion.

@@

\subsection{Unification For Strictly Curried Functions}
Although we omit most details of unification, we shall discuss it for curried
functions, as it is essential in the extended elaboration algorithm. We seek to
solve meta-telescopes in a way such that iterated implicit function types on the
two sides of a unification problem become definitionally equal. First we define
a helper function which counts leading implicit function types:
\begin{alignat*}{3}
  & \implArity\,(\{x : A\}\to B) && :\equiv 1 + \implArity\,B\\
  & \implArity\,A                && :\equiv 0
\end{alignat*}
Then, we extend unification. The most interesting case is when we unify a
curried function type with an implicit function type. In this case, if there are
``missing'' implicit arguments on the side of the curried function type (note
the $\implArity$ condition on the pattern), we refine the $A$ telescope domain
to an extended $(x_0 : A_0) \TCons A_1$ telescope. Since we invent a fresh $A_1$
domain for a curried function type, we need to add a constancy constraint for it
as well.
\begin{alignat*}{3}
& \unify_{\Theta_0|\Gamma}(\{x : \ol{A}\}\to B)\,(\{x_0 : A_0\}\to B')\,\,|\,\,\implArity\,B < \implArity\,B' + 1 :\equiv \edo\\
& \quad \slet\,(\Theta_1,\,\theta_1,\,A_1) = \freshMeta{\Theta_0}{\Gamma,\,x_0 : A_0}{\Tel}\\
& \quad (\Theta_2,\,\theta_2) \leftarrow
  \unify_{\Theta_1|\Gamma[\theta_1]}\,(A[\theta_1])\,((x : A_0[\theta_1])\TCons A_1)\\
& \quad \slet\,(\Theta_3,\,\theta_3) =
    \newConstancy{\Theta_2}{\Gamma[\theta_{12}],\,x_0 : A_0[\theta_{12}]}{x_1}{(A_1[\theta_2])}{(B[\theta_{12}][x\mapsto (x_0 :: x_1)])}\\
& \quad\unify_{\Theta_3|\Gamma[\theta_{123}],\,x_0 : A_0[\theta_{123}]}\,
         (\{x_1 : A_1[\theta_{23}]\}\to B[\theta_{123}][x\mapsto (x_0 :: x_1)])\,B'
\end{alignat*}
We have the symmetric $\unify_{\Theta_0|\Gamma}(\{x_0 : A_0\}\to B')\,(\{x :
\ol{A}\}\to B)$ case the same way as above.

Below, let us assume that $B'$ is not meta-headed, and it also does not match
the pattern above. Hence, either $B'$ is not an implicit function type, or the
arity condition does not hold, so we learn that the domain telescope should be
empty.
\begin{alignat*}{3}
& \unify_{\Theta_0|\Gamma}(\{x : \ol{A}\}\to B)\,B' :\equiv \edo\\
& \quad (\Theta_1,\,\theta_1) \leftarrow \unify_{\Theta_0|\Gamma}\,A\,\epsilon\\
& \quad \unify_{\Theta_1|\Gamma[\theta_1]}\,(B[\theta_1][x\mapsto[]])\,(B'[\theta_1])
\end{alignat*}
Again, we also have the symmetric case. Some examples for unification behavior:
\begin{itemize}
\item If $B$ is not meta-headed, $\unify\,(\{x : \ol{\alpha}\}\to B)\,B$ causes
  $\alpha$ to be solved as $\epsilon$, for the following reason: if $B$ is an
  implicit function type, the arity condition fails, otherwise the pattern match
  fails, so either way we fall through the first rule and solve $\alpha$ to
  $\epsilon$.
\item If $B$ is again not meta-headed, $\unify\,(\{x : \ol{\alpha}\}\to B)\,(\{x
  : A\}\to B)$ causes $\alpha$ to be solved as $A \TCons \epsilon$.
\end{itemize}

For telescopes, $\Rec$, inhabitants of $\Rec$, curried abstraction and
application, unification is structural, and other cases remain the same as in
the basic elaborator of Section \ref{sec:bidirectional_elaboration}.

@@

In the definition of checking, we insert a new clause after
$\echeck{t}{\Theta}{\Gamma}{(\{x : A\}\to B)}$:
\begin{alignat*}{3}
& \echeck{t}{\Theta_0}{\Gamma}{(\alpha\,\ol{u})} :\equiv \edo\\
& \quad \slet\,(\Theta_1,\,\theta_1,\,A) = \freshMeta{\Theta_0}{\Gamma}{\Tel}\\
&\quad(\Theta_2,\,\theta_2,\,t',\,B)\leftarrow
            \einsert\,(\einfer{t}{\Theta_1}{\Gamma[\theta_1],\,x:\Rec\,A}) \\
& \quad \slet\,(\Theta_3,\,\theta_3) = \newConstancy{\Theta_2}{\Gamma[\theta_{12}]}{x}{(A[\theta_2])}{B}\\
& \quad (\Theta_4,\,\theta_4) \leftarrow
            \unify\, ((\alpha\,\ol{u})[\theta_{123}])\,(\{x : \ol{A[\theta_{23}]}\} \to B[\theta_3])\\
& \quad \ereturn\,(\Theta_4,\,\theta_{1234},\,(\lambda\,\{x : \ol{A[\theta_{234}]}\}.\,t'[\theta_{34}]))
\end{alignat*}
Hence, when checking a term with a meta-headed type, we create a fresh meta with
$\Tel$ type, infer a type for the term (avoiding superfluous insertion, as in
Section \ref{sec:elaboration} when inferring type for $\lambda$-s) and wrap the
result in a strictly curried $\lambda$. We again need to create new constancy
constraint. This way, if $B$ does not depend on $x : \Rec\,A$, the strictly
curried $\lambda$ in the output immediately computes away, since $A$ is solved
to $\epsilon$.

@@

On line $\scriptstyle{14}$ we have the case when a curried function type is
matched with a type which is not a curried function, implicit function or a
meta-headed type. We accordingly unify domain telescope $\alpha_3\,A$ with
$\epsilon$, which causes $\alpha_3$ to be solved as $\lambda\,A.\,\epsilon$
according to standard pattern unification. Now, $\alpha_2\,\{A\}\,\{z :
\ol{\alpha_3\,A}\}$ computes to $\alpha_2\,\{A\}\,\{z : \epsilon\}$, which
further computes to $\alpha_2\,\{A\}$ by \LabTirName{app-$\epsilon$}. From here,
unification finishes with success. On the last line, the expected output is
returned: since the $\alpha_1$ telescope meta is now solved, curried function
types and abstractions are computed away.

Computing
$\echeckt{\mi{cons}\,(\lambda\,\{A\}\,x.\,x)\,\mi{nil}}{(\mi{List}\,(\{A :
  \U\}\to A \to A))}$ also succeeds, as in Agda and our basic elaborator. In
this case elaboration also inserts a curried $\lambda$ around
$(\lambda\,\{A\}\,x.\,x)$, but its domain is then solved to $\epsilon$ according
to the rules for curried function type unification.

\end{example}
